\Sof{TBA - New Titles in the comments}

### 2020-05-25 paper/icsme20.tex
intro/abstract

\section{Introduction}
%
Software quality is important because it is ultimately related to the overall
cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. 
Software quality characteristics include, but are not limited to functional correctness,
reliability, usability, maintainability, and security. Security is an 
important non-functional requirement during the development of software systems. 
However, there is still a considerable amount of vulnerabilities being
discovered, and fixed, almost weekly as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available at
\url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
These vulnerabilities are gateways for attackers to penetrate systems and 
exploit their resources. As an example, at the beginning of $2020$, academics published a 
new execution attack (CacheOut) capable of leaking data from Intel CPUs across many 
security boundaries~\cite{cacheOut} $-$ a processor bug that the company attempted to 
address in previous generations and that makes every user using a CPU released before Q$4$ $2018$ vulnerable. Although not a software bug, software can mitigate this type of issues
at the cost of features and/or performance.
%

In $2011$, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010 considering
\emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. 
However, many developers still lack the knowledge about best practices to deliver secure and high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
supports the production of more reliable software preventing the introduction of 
faults/vulnerabilities in the code bases. ISO describes software 
maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010~\cite{iso:2011}. Novel tools have been built to 
detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
Despite these efforts, improving software security is not a trivial task and requires 
implementing patches that might affect software maintainability. 
Our suspiciousness is that some of the these patches may have a negative 
impact on the software maintainability and, possibly, even be the cause of the 
introduction of new vulnerabilities $-$ harming software reliability. Therefore, 
in this paper, we present an empirical study on the impact of patches of 
vulnerabilities on software maintenance across open-source software.
%
% Software maintainability can also be defined as 
% the property of software that reveals how easily the software system can be 
% maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure. 
% Hedegus et al.
% ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
% groups of refactored elements and non-refactored and conclude that the source
% code elements subjected to refactorings had significantly lower maintainability
% than the elements not modified.
% There is still little information about the impact of patching vulnerabilities
% in open-source software maintainability. \textit{Are developers affecting software
% maintainability when trying to improve software security?}, it is what we want to
% investigate.
%

As ISO does not provide any specific guidelines/formulas to calculate 
maintainability, we resort to Software Improvement Group (SIG)'s web-based source 
code analysis service Better Code Hub (BCH) to compute the software compliance 
with a set of $10$ guidelines/metrics to produce quality software 
based in ISO/IEC $25010$~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}. 
There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical evaluation;
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one of the
first steps taken by SIG to introduce security on their maintainability model). 
%
We use BCH's toolset to provide an assessment of maintainability in 
software mainly due to the broad number of metrics used.

From a methodological 
perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
software. We calculate software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes 
applied in the code bases to patch vulnerabilities affect code maintainability.
Results also show that different levels of attention should be given to different
severity levels, programming languages and types of weaknesses.
Little information is known about the impact of security patches on the software
maintainability. With this study, 
we intend to highlight the need for tools to predict the impact of patches on 
maintainability~\cite{4724577}, documentation to help developers adopt security 
best practices~\cite{6311252, 7927935, MESQUIDA201519} and the importance of integrating
security-related code evolution in computer curricula.
%
%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
%

>>>>

\section{Introduction}
%
Software quality is important because it is ultimately related to the overall
cost of developing, extending and maintaining software applications~\cite{slaughter1998evaluating}. 
Software quality characteristics include, but are not limited to functional correctness,
reliability, usability, maintainability, and security. Security is an 
important non-functional requirement during the development of software systems. 
However, there is still a considerable amount of vulnerabilities being
discovered and fixed, almost weekly, as disclosed by the Zero Day Initiative
website\footnote{Zero Day Initiative website available at
\url{https://www.zerodayinitiative.com/advisories/published/} (Accessed on \today{})}.
%These vulnerabilities are gateways for attackers to penetrate systems and 
%exploit their resources. As an example, at the beginning of $2020$, academics published a 
%new execution attack (CacheOut) capable of leaking data from Intel CPUs across many 
%security boundaries~\cite{cacheOut} $-$ a processor bug that the company attempted to 
%address in previous generations and that makes every user using a CPU released before Q$4$ 
%$2018$ vulnerable. Although not a software bug, software can mitigate this type of issues
%at the cost of features and/or performance.
%

In $2011$, the International Organization for Standardization (ISO) issued an
update for software product quality ISO/IEC 25010 considering
\emph{Security} as one of the main software product quality characteristics~\cite{iso:2011}. 
However, many developers still lack the knowledge about best practices to deliver secure and 
high-quality software~\cite{Pothamsetty:2005:SEL:1107622.1107635, 8077802}.
\Rui{link between security and maintainability is still obscure.}
Unmaintainable code is hard to test, analyze and re-use, and thus more difficult
to find vulnerabilities and defects. Therefore, writing maintainable code
supports the production of more reliable software preventing the introduction of 
faults/vulnerabilities in the code bases. ISO describes software 
maintainability as ``the degree of effectiveness and
efficiency with which a software product or system can be modified to improve
it, correct it or adapt it to changes in environment, and in requirements'' on
software quality ISO/IEC 25010~\cite{iso:2011}. Novel tools have been built to 
detect automatically software vulnerabilities (e.g., FindBugs, Infer and more). 
Despite these efforts, improving software security is not a trivial task and requires 
implementing patches that might affect software maintainability. 
Our \Rui{wrong word -->} suspiciousness is that some of the these patches may have a negative 
impact on the software maintainability and, possibly, even be the cause of the 
introduction of new vulnerabilities $-$ harming software reliability. Therefore, 
in this paper, we present an empirical study on the impact of patches of 
vulnerabilities on software maintenance across open-source software.
%
% Software maintainability can also be defined as 
% the property of software that reveals how easily the software system can be 
% maintained~\cite{IEEEComputerSociety:2014:GSE:2616205} or recover after a failure. 
% Hedegus et al.
% ($2018$)~\cite{HEGEDUS2018313}, analyzed the difference of maintainability between
% groups of refactored elements and non-refactored and conclude that the source
% code elements subjected to refactorings had significantly lower maintainability
% than the elements not modified.
% There is still little information about the impact of patching vulnerabilities
% in open-source software maintainability. \textit{Are developers affecting software
% maintainability when trying to improve software security?}, it is what we want to
% investigate.
%

As ISO does not provide any specific guidelines/formulas to calculate 
maintainability, we resort to Software Improvement Group (SIG)'s web-based source 
code analysis service Better Code Hub (BCH) to compute the software compliance 
with a set of $10$ guidelines/metrics to produce quality software 
based in ISO/IEC $25010$~\cite{Visser:2016:OREILLY}. The code metrics used by BCH 
were empirically validated in previous work~\cite{Bijlsma:2012:FIR:2317098.2317124, 8530041, cruz2019energyoriented}. 
There are other well-known standards and models that have been proposed
to increase software security: Common Criteria~\cite{common:2009} which received
negative criticism regarding the costs associated and poor technical evaluation;
the OWASP Application Security Verification Criteria~\cite{oswap:2009} which is
focused only on web applications, and a model proposed by Xu et al. ($2013$)
~\cite{6616351} for rating software security (arguably, it was one of the
first steps taken by SIG to introduce security on their maintainability model). 
%
We use BCH's toolset to provide an assessment of maintainability in 
software mainly due to the broad number of metrics used. \Rui{this paragraph is boring and in cicles. revise.}

From a methodological 
perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
software. We calculate software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes 
applied in the code bases to patch vulnerabilities affect code maintainability.
\Rui{Sentence is too criptic -->} Results also show that different levels of attention should be given to different
severity levels, programming languages and types of weaknesses.
Little information is known about the impact of security patches on the software
maintainability. With this study, 
we intend to highlight the need for tools to predict the impact of patches on 
maintainability~\cite{4724577}, documentation to help developers adopt security 
best practices~\cite{6311252, 7927935, MESQUIDA201519} and the importance of integrating
security-related code evolution in computer curricula.
%
%BCH compiles the ISO/IEC 25010 into a set of $10$ guidelines that are the
%base of BCH's evaluation. Two examples are the \emph{Write Simple Units of Code}
%based on the McCabe Complexity~\cite{1702388} and \emph{Keep your code base Small}
%based on the idea that smaller code bases are easier to maintain and less prone to defects~\cite{Visser:2016:OREILLY}.
%
